---
title: "TRD"
output: html_document
date: "2025-09-18"
---

<style type="text/css">
.main-container {
max-width: 80% !important;
margin: auto;
}
</style>

```{r, include=FALSE}
source("../R/my_functions.R")
notebook_setup()

options(scipen = 999)
knitr::opts_chunk$set(warning = FALSE)
```

# Load in the Medication table
```{r}
folder_date <- "25_09_10"
file_date <- "25_09_10"
my_version <- "v1"

load(paste0("../PrepData-", folder_date, "/Medication_Table-", file_date, "-", my_version, ".rds"))
medi_table_prep <- medi_table_prep %>% dplyr::select(-EncounterId_SH)
medi_table_prep %>%
  my_kable_table()
```

# Frequency of Pharmaceutical Classes in the dataset
```{r}
medi_table_prep$MedPharmaceuticalClass %>% 
  table() %>% 
  as.data.frame %>% 
  rename(Drug = ".") %>% 
  arrange(-Freq) %>% 
  mutate(Freq = comma(Freq)) %>% 
  my_kable_table(cut = F)
```

# Filter to antidepressants
```{r}
ad_record <- medi_table_prep %>% filter(MedPharmaceuticalClass == "Antidepressants") %>% mutate(across(where(is.factor), ~as.character(.)))
write.table(unique(ad_record$PatientEpicId_SH), paste0("../OutputData-", folder_date, "/IDs-Antidepressant_Use.txt"), row.names = F, col.names = F)
ad_record %>%
  my_kable_table()
```

# Frequency of Order Types in the Antidepressants Dataset.
```{r}
ad_record$MedOrderType %>% table() %>% as.data.frame %>% rename(Drug = ".") %>% arrange(-Freq) %>% mutate(Freq = comma(Freq)) %>% 
  my_kable_table()
```

# Frequency of Medication Names in the Antidepressants Dataset.
```{r}
ad_record$MedSimpleGenericName %>% table() %>% as.data.frame %>% rename(Drug = ".") %>% arrange(-Freq) %>% mutate(Freq = comma(Freq)) %>% 
  my_kable_table(cut = F)
```

# Filter the Diagnosis Table to MDD
```{r}
# mdd_codes <- c("296.2", "296.3", "300.4", "311", "F32", "F33", "F34.1")
load(paste0("../PrepData-", folder_date, "/MDD-Diagnosis_Table-", file_date, "-", my_version, ".rds"))
mdd_record <- diag_record
load(paste0("../PrepData-", folder_date, "/MDD_IDs-", file_date, "-", my_version, ".rds"))
mdd_id <- data.frame(PatientEpicId_SH = diag_id)
mdd_record %>%
  my_kable_table()
```

# Define Bipolar Disorder and Schizophrenic Participants.
```{r}
# bd_codes <- c("296.0", "296.4", "296.5", "296.6", "296.7", "296.8", "F30", "F31")
load(paste0("../PrepData-", folder_date, "/BD_IDs-", file_date, "-", my_version, ".rds"))
bd_id <- data.frame(PatientEpicId_SH = diag_id)

# sch_codes <- c("295", "298", "F20", "F23", "F25", "F28", "F29")
load(paste0("../PrepData-", folder_date, "/SCH_IDs-", file_date, "-", my_version, ".rds"))
sch_id <- data.frame(PatientEpicId_SH = diag_id)

cat(paste0("N BD: ",(bd_id %>% nrow %>% comma), "\n"))
cat(paste0("N SCH: ",(sch_id %>% nrow %>% comma)))
```

# Identify First MDD Record.
```{r}
mdd_first_record <- mdd_record %>% 
  group_by(PatientEpicId_SH) %>% 
  filter(Encounter_StartVisit == min(Encounter_StartVisit)) %>% 
  dplyr::select(PatientEpicId_SH, Encounter_StartVisit) %>%
  rename(first_depression_dx_date = Encounter_StartVisit) %>%
  slice(1) %>%
  ungroup()

mdd_first_record %>% 
  my_kable_table()
```

# Total MDD in Dataset.
```{r}
paste0("There are ",(mdd_first_record %>% nrow %>% comma)," participants in the dataset with Depression.") %>% cat
```

```{r}
cat(paste0("There are ",  (mdd_id %>% filter(!PatientEpicId_SH %in% c(bd_id$PatientEpicId_SH, sch_id$PatientEpicId_SH)) %>% nrow %>% comma), " participants that are MDD and do not have Schizophrenia or Bipolar Disorder"))
```

# Filter to Antidepressants that are prescribed after first MDD Diagnosis.
```{r}
ad_record_1 <- ad_record  %>%
  filter(PatientEpicId_SH %in% unique(mdd_first_record$PatientEpicId_SH)) %>%
  left_join(mdd_first_record, by = join_by(PatientEpicId_SH == PatientEpicId_SH)) %>% 
  filter(MedStartInstant >= first_depression_dx_date)


paste0("There are ",(ad_record$PatientEpicId_SH %>% unique %>% length %>%comma)," participants in the antidepressant drug dataset.\nThe dataset has ",(ad_record %>% nrow %>% comma)," antidepressant drug records.") %>% cat

paste0("\n\nThere are ",(ad_record_1$PatientEpicId_SH %>% unique %>% length %>%comma)," participants in the antidepressant drug dataset whose MDD diagnosis precedes antidepressant treatment.\nThe dataset filtered to post-MDD treatment has ",(ad_record_1 %>% nrow %>% comma)," antidepressant drug records.") %>% cat
```

# Identify first ad prescription post MDD diagnosis
```{r}
post_mdd_ad_index <- ad_record_1 %>%
  arrange(PatientEpicId_SH, MedStartInstant) %>% 
  group_by(PatientEpicId_SH) %>%
  filter(MedStartInstant == min(MedStartInstant)) %>%
  mutate(n_drugs_prescribed = n(),
         n_unique_drugs_prescribed = length(unique(MedSimpleGenericName))
         ) 

filename <- paste0("../OutputData-", folder_date, "/post_mdd_ad_index.csv")
post_mdd_ad_index %>% write.csv(filename, row.names = F)

post_mdd_ad_index %>%
  my_kable_table()
```

# Create a list of all drugs that each participant has taken.
```{r}
antidepressant_record_uq <- ad_record_1 %>% 
  dplyr::select(PatientEpicId_SH, MedSimpleGenericName) %>% 
  unique %>%
  arrange(PatientEpicId_SH)

antidepressant_record_uq %>%
  my_kable_table()
```

# TRD definition step 1
```{r}
ad_record_1$days_supply <- NA
filename <- paste0("../OutputData-", folder_date, "/antidepressant_consecutive_instance.csv")
overwrite <- TRUE
if(!file.exists(filename) | overwrite){
  df <- ad_record_1
  
  # Calculate intervals and instances
  df <- df %>%
    arrange(PatientEpicId_SH, MedSimpleGenericName, MedStartInstant) %>%
    group_by(PatientEpicId_SH, MedSimpleGenericName) %>%
    mutate(
      interval = as.numeric(difftime(MedStartInstant, lag(MedStartInstant), units = "days")),
      interval = ifelse(row_number() == 1, 1, interval),
      consecutive_instance = interval > 180,
      consecutive_instance = ifelse(row_number() == 1, TRUE, consecutive_instance),
      consecutive_instance = cumsum(consecutive_instance)
    )
  
  # Calculate first and last records
  consecutive_instance_tab <- df %>%
    group_by(PatientEpicId_SH, MedSimpleGenericName, consecutive_instance) %>%
    summarise(
      first.record = first(MedStartInstant),
      last.record = last(MedStartInstant),
      last.record.days_supply = last(days_supply),
      drug_class = first(MedPharmaceuticalSubclass),
      .groups = 'drop'
    ) %>%
    dplyr::select(-consecutive_instance) %>%
    arrange(PatientEpicId_SH, first.record, MedSimpleGenericName) %>%
    mutate(last.record = replace(last.record, last.record == first.record, as.Date(NA))) %>%
    as.data.frame 
  consecutive_instance_tab %>% write.csv(filename, row.names = FALSE)
}else{
  consecutive_instance_tab <- read.csv(filename)
}

consecutive_instance_tab %>% 
  my_kable_table()
```

# TRD definition step 2.
```{r}
overwrite <- TRUE
my_progress <- 0
filename <- paste0("../OutputData-", folder_date, "/antidepressant_consecutive_period.csv")
if(!file.exists(filename) | overwrite){
  assign_period <- function(this_data){
    for(i in 1:nrow(this_data)){
      this_period_time_from_start <- ifelse(i == 1, 
                                            0, 
                                            as.integer(difftime(as_date(this_data$first.record[i]), 
                                                                as_date(this_data$period_start[i-1]), 
                                                                units = "days"))
      )
      
      new_period <- ifelse(i == 1, 
                           TRUE, 
                           (this_period_time_from_start > 365 | (this_data$MedSimpleGenericName[i] == this_data$MedSimpleGenericName[i-1]))
      )
      
      this_data$period[i] = ifelse(i == 1, 
                                   1, 
                                   ifelse(new_period,
                                          this_data$period[i-1] + 1,
                                          this_data$period[i-1]
                                   )
      )
      
      this_data$period_start[i] <- ifelse(new_period, 
                                          this_data$first.record[i], 
                                          this_data$period_start[i-1])
      
      this_data$period_time_from_start[i] <- ifelse(new_period, 
                                                    0, 
                                                    this_period_time_from_start)                       
    }
    
    return(this_data)
  }
  
  consecutive_period_tab_start <- consecutive_instance_tab %>%
    mutate(
      period = NA,
      period_start = as.Date(NA),
      period_time_from_start = NA
    ) %>%
    split(consecutive_instance_tab$PatientEpicId_SH)
  
  # Register the parallel backend
  num_cores <- detectCores()
  cl <- makeCluster(num_cores/2)
  registerDoParallel(cl)
  
  # Use foreach to apply your function in parallel
  consecutive_period_tab_list <- foreach(x=consecutive_period_tab_start, .packages = c("lubridate","base")) %dopar% {
    assign_period(x)
  }
  
  # Stop the cluster when you're done
  stopCluster(cl)
  
  consecutive_period_tab <- bind_rows(consecutive_period_tab_list)
  consecutive_period_tab %>% write.csv(filename, row.names = F)
}else{
  consecutive_period_tab <- read.csv(filename)
}

consecutive_period_tab %>%
  my_kable_table()
```

```{r}
third_drug_of_trd_index <- consecutive_period_tab %>%
  group_by(PatientEpicId_SH, period, MedSimpleGenericName) %>%
  slice(1) %>%
  group_by(PatientEpicId_SH, period) %>%
  mutate(ndrug = n(),
         drug_order = row_number()) %>% 
  ungroup %>%
  filter(ndrug > 2) %>% 
  filter(!PatientEpicId_SH %in% c(bd_id$PatientEpicId_SH, sch_id$PatientEpicId_SH)) %>% 
  group_by(PatientEpicId_SH) %>% 
  filter(period == min(period)) %>% 
  ungroup %>% 
  filter(drug_order == 3)

filename <- paste0("../OutputData-", folder_date, "/TRD_index_period_third_drug.csv")
third_drug_of_trd_index %>% write.csv(filename, row.names = F)

third_drug_of_trd_index %>%
  my_kable_table()
```

# TRD definition step 3.
```{r}
consecutive_period_tab_summ <- consecutive_period_tab %>% group_by(PatientEpicId_SH,period) %>% 
  summarise(
    ndrug = (MedSimpleGenericName %>% unique %>% length),
    drugs = (MedSimpleGenericName %>% unique %>% paste(collapse = ", ")),
    period.start = min(first.record), period.end = max(first.record))
consecutive_period_tab_summ %>% head(10)
write.csv(consecutive_period_tab_summ, paste0("../OutputData-", folder_date, "/antidepressant_consecutive_period_tab_summ.csv"))
consecutive_period_tab_summ %>% 
  my_kable_table()
```

# TRD index definition step 4.
```{r}
consecutive_period_trdIndexEvent <- consecutive_period_tab_summ %>% 
  filter(ndrug > 2) %>% group_by(PatientEpicId_SH) %>% filter(period == min(period)) %>% 
  filter(!PatientEpicId_SH %in% c(bd_id$PatientEpicId_SH, sch_id$PatientEpicId_SH))
write.csv(consecutive_period_trdIndexEvent, paste0("../OutputData-", folder_date, "/TRD_index.csv"))
consecutive_period_trdIndexEvent %>%
  my_kable_table()
```

# TRD legacy step 4.
```{r}
consecutive_period_maxDrugs <- consecutive_period_tab_summ %>% group_by(PatientEpicId_SH) %>% 
  filter(ndrug == max(ndrug)) %>% filter(period == max(period))
consecutive_period_maxDrugs$has_TRD <- consecutive_period_maxDrugs$ndrug > 2
write.csv(consecutive_period_maxDrugs, paste0("../OutputData-", folder_date, "/antidepressant_consecutive_period_maxDrugs.csv"))
consecutive_period_maxDrugs %>%
  my_kable_table()
```

# Frequency of the number of max drugs in a period.
```{r}
consecutive_period_maxDrugs$ndrug %>% table %>% as.data.frame %>% rename(nDrugs = ".") %>% arrange(-Freq) %>% mutate(Freq = comma(Freq)) %>% 
  my_kable_table()
```

# TRD count.
```{r}
TRD_list <- consecutive_period_trdIndexEvent %>% 
  dplyr::select(PatientEpicId_SH) %>% 
  filter(!PatientEpicId_SH %in% c(bd_id$PatientEpicId_SH, sch_id$PatientEpicId_SH))

cat(paste0("N TRD: ", TRD_list %>% nrow %>% comma, "\n"))
```

# Write the TRD data.
```{r}
write.table(unique(TRD_list$PatientEpicId_SH), paste0("../OutputData-", folder_date, "/IDs-TRD.txt"), row.names = F, col.names = F)
```